---
title: "Info Visualisation Notebook"
output: info_vis_notebook
---

Set working directory

```{r}
# Ana's WD
# setwd("C:/Users/Ana-Maria's Laptop/Documents/GitHub/infoVisualization")

# Andrei's WD
# setwd("C:/Users/andre/Documents/infoVisualization/")
```

Imports - libraries and read dataset

```{r}
#install.packages("png")
#install.packages("gifsky")
#install.packages("gganimate")
#install.packages("gapminder")
#install.packages("ggmap")
#install.packages("esquisse")
#install.packages("ggiraph")
#install.packages("gdtools")
#install.packages("leaflet") # interactive maps
#install.packages("plotly")




library(ggplot2)
library(gganimate)
library(gapminder)
#library(tidyverse)
library(ggmap)
library(ggiraph)

library(esquisse)
library(leaflet)
library(plotly)


# esquisser(df)
```

```{r}
earthquakes_csv = "eartquakes_Romania.csv"
df <- read.csv(file = earthquakes_csv)
```

```{r}
# convert time column to R timestap
df$time <- gsub('T', ' ', df$time)
df$time <- gsub('Z', '', df$time)
df$time <- as.POSIXct(df$time, format="%Y-%m-%d %H:%M:%OS", tz=Sys.timezone())

#split timestamp into date and time
df$timeYear <- as.Date(df$time)
df$timeDay <- format(df$time, "%H:%M:%OS")

df$year <- format(df$time, "%Y")
df$month <- format(df$time, "%m")

df$yearInt <- as.integer(df$year)
```

```{r}
# see data types
str(df)
```
Describing the dataset:
time - self-explanatory, includes: year, month, day, hour, minute, seconds, milliseconds
latitude & longitude - the coordinates of the earthquake source
depth - the depth at which it originated
mag - the physical size of the earthquake
  Great	8 or more
  Major	7 - 7.9
  Strong	6 - 6.9
  Moderate	5 - 5.9
  Light	4 - 4.9
  Minor	3 - 3.9
magType - category based on magnitude range and distance range. Details: https://www.usgs.gov/natural-hazards/earthquake-hazards/science/magnitude-types?qt-science_center_objects=0#qt-science_center_objects

nst - number of seismic stations which reported P- and S- arrival times for the earthquake
gap - azimuthal gap range (range: 0 to 360)
dmin - distance from the epicenter to the nearest station. In general, the smaller this number, the more reliable the calculated depth of the earthquake
rms - root-mean-squared residual of solution (range: 0. to 1.)
net - ?
id 
updated
place - explained in km and cardinal direction from nearest relevant settlement
type - if it is an earthquake or something else. All of them are the same type, we ignore this column
horizontalError
depthError
magError
magNst - number of stations that reported the magnitude 
status: all of them are reviewed
locationSource - the source of the earthquake location data (the data is taken from multiple websites)
magSource - the source of the earthquake magnitude data (the data is taken from multiple websites)
Extracted columns
timeYear - the date at which it occured (year, month and day)
timeDay - the time at which it occured (hour, minutes, seconds, milliseconds)
year
month
yearInt - same as year, but as int datatype, in order to use it for animated graphs




We start by exploring the variance in the dataset. This way we can better understand each column, besides its initial definition.

```{r}
mag_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = mag), fill = "steelblue", bins = 40) +
  theme_minimal()
mag_hist + ggtitle("Number of recorded earthquakes by magnitude")

year_bar <- ggplot(df) + 
  geom_bar(mapping = aes(y = year), fill = "steelblue") +
  theme_minimal()
year_bar + ggtitle("Number of earthquakes per year (1990 - 2017)")

month_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = month), fill = "steelblue") +
  theme_minimal()
month_bar + ggtitle("Number of earthquakes by month (1990 - 2017)")

depth_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = depth), fill = "steelblue") +
  theme_minimal()
depth_bar + ggtitle("Number of earthquakes by depth")

magtype_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = magType), fill = "steelblue") +
  theme_minimal()
magtype_bar + ggtitle("Magnitude type (category)")

nst_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = nst), fill = "steelblue", bins=50) +
  theme_minimal()
nst_bar + ggtitle("Nst - number of seismic stations reporting")

gap_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = gap), fill = "steelblue") +
  theme_minimal()
gap_bar + ggtitle("Gap Value (azimuthal gap range)")

dmin_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = dmin), fill = "steelblue") +
  theme_minimal()
dmin_bar + ggtitle("Dmin Value (distance from epicenter)")

rms_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = rms), fill = "steelblue") +
  theme_minimal()
rms_bar + ggtitle("Rms Value (root-mean-squared residual of solution)")

net_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = net), fill = "steelblue") +
  theme_minimal()
net_bar + ggtitle("Net Value")

type_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = type), fill = "steelblue") +
  theme_minimal()
type_bar + ggtitle("Type of data entry")

status_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = status), fill = "steelblue") +
  theme_minimal()
status_bar + ggtitle("Status of data entry")

locationSource_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = locationSource), fill = "steelblue") +
  theme_minimal()
locationSource_bar + ggtitle("Location: data source")

magSourceSource_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = magSource), fill = "steelblue") +
  theme_minimal()
magSourceSource_bar + ggtitle("Magnitude: data source")

horizontalError_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = horizontalError), fill = "steelblue") +
  theme_minimal()
horizontalError_hist + ggtitle("Horizontal Error")

depthError_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = depthError), fill = "steelblue") +
  theme_minimal()
depthError_hist + ggtitle("Depth Error ")

magError_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = magError), fill = "steelblue") +
  theme_minimal()
magError_hist + ggtitle("Magnitude Error")

magNst_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = magNst), fill = "steelblue") +
  theme_minimal()
magNst_hist + ggtitle("Number of stations that reported the magnitude ")
```

We start by exploring the most obvious questions: where do most earthquakes happen, including their magnitude as well


```{r}
ggplot(data=df) +
  geom_point(mapping = aes(x = longitude, y = latitude, color=mag))
```

As we can see, there is a big cluster where most earthquakes happen, with a smaller cluster that is noticeable, followed by sparsely distributed cases
Let's now plot them on a map for a better visualization. Based on common knowledge of earthquakes in Romania, can you guess where the main epicenter will be?

(we are using stamen source for the map, but the commented code also uses google maps. It requires an API key, though)

```{r}
romania_map <- get_map("Brasov", maptype = "roadmap", zoom = 6)

romania_box <- c(20, 43, 30, 48.5)
romania_map2 <- get_map(location = romania_box, source = "stamen", maptype = "toner", zoom = 7)


#ggmap(romania_map) + 
#  geom_point(data = df, aes(x = longitude, y = latitude), color = "steelblue", size = 1) 

ggmap(romania_map2) + 
  geom_point(data = df, aes(x = longitude, y = latitude, color = mag), size = 1) 


```
As we can see, most earthquakes are in Vrance County, as expected.

Let's also make an interactive map to play around it

We have colored each pin by earthquake classification type (based on magnitude) and have grouped them in clusters (and an extra without clusters just for showcase).

green: minor
blue: light
pink: moderate
orange: strong
red: major
black: great

As we can see, most of the recorder earthquakes are either minor or light. As we go up in magnitude, the number of earthquakes decreases.
```{r}
getColor <- function(quakes) {
  sapply(quakes$mag, function(mag) {
  if (mag < 4) {
    "green"
  } else if (mag < 5) {
    "blue"
  } else if (mag < 6) {
    "pink"
  } else if (mag < 7) {
    "orange"
  } else if (mag < 8) {
    "red"
  } else {
    "black"
  } })
}

icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'black',
  library = 'ion',
  markerColor = getColor(df)
)

leaflet(data = df) %>% 
  addTiles() %>%
  addAwesomeMarkers(~longitude, ~latitude, popup = ~as.character(year), label = ~as.character(mag), icon=icons, clusterOptions = markerClusterOptions())


leaflet(data = df) %>% 
  addTiles() %>%
  addAwesomeMarkers(~longitude, ~latitude, popup = ~as.character(year), label = ~as.character(mag), icon=icons)
```


Let's start exploring covariance, as well - starting with the number of earthquakes per magnitude, each year.
We started by using a barplot and a freqpoly, but it is quite hard to get useful information due to different number of earthquakes in each year. Thus, we switched to density plot which takes that into consideration, as well.

```{r}
ggplot(data=df) +
  geom_bar(mapping = aes(x = mag, fill = year))

ggplot(data=df) +
  geom_freqpoly(mapping = aes(x = mag, color = year))

ggplot(data=df) +
  geom_density(mapping = aes(x = mag, color = year))
```

Due to the number of years, it is still hard to read the graphs. Let's use a facet_wrap for a better understanding.

See? Way better.

```{r}
ggplot(data=df) +
  geom_bar(mapping = aes(x = mag, fill = year)) +
  facet_wrap(~ year) +
  theme_minimal()

ggplot(data=df) +
  geom_freqpoly(mapping = aes(x = mag, color = year)) +
  facet_wrap(~ year) +
  theme_minimal()

ggplot(data=df) +
  geom_density(mapping = aes(x = mag, color = year)) +
  facet_wrap(~ year) +
  theme_minimal()

```
There is obviously a correlation between the location and the number of earthquakes. Let's see if we can find other correlations, as well. How about between the depth and the magnitude

```{r}

ggplot(data=df) +
  geom_point(mapping = aes(x = depth, y = mag, color = year))


ggplot(data=df) +
  geom_point(mapping = aes(x = depth, y = mag, color = year)) +
  facet_wrap(~ year) +
  theme_minimal()


ggplot(data=df) +
  geom_line(mapping = aes(x = depth, y = mag, color = year)) +
  facet_wrap(~ year) +
  theme_minimal()

```

Ok, let's swithc to 3D plots as it seems we need to take into account more than 2 columns.

Starting with longitude, latitude and magnitude.

```{r}

longlatmag <- plot_ly(df, x = ~longitude, y = ~latitude, z = ~mag,
               marker = list(color = ~mag, colorscale = c('#FFE1A1', '#683531'), showscale = TRUE))
longlatmag <- fig %>% add_markers()
longlatmag <- fig %>% layout(scene = list(xaxis = list(title = 'Longitude'),
                                   yaxis = list(title = 'Latitude'),
                                   zaxis = list(title = 'Magnitude')),
                      annotations = list(
                        x = 1.13,
                        y = 1.05,
                        text = 'Magnitude',
                        xref = 'mag',
                        yref = 'mag',
                        showarrow = FALSE
                        ))
longlatmag
```

Ok, now let's take a look at the latitude, longitude and depth.

It is interesting to see that while there are 2 obvious clusters regarding the location, one cluster (Vrancea one) definitely has earthquakes with more depth.
We have also colored them based on magnitude.


```{r}

longlatdepth <- plot_ly(df, x = ~longitude, y = ~latitude, z = ~depth,
               marker = list(color = ~mag, colorscale = c('#FFE1A1', '#683531'), showscale = TRUE))
longlatdepth <- fig %>% add_markers()
longlatdepth <- fig %>% layout(scene = list(xaxis = list(title = 'Longitude'),
                                   yaxis = list(title = 'Latitude'),
                                   zaxis = list(title = 'Depth')),
                      annotations = list(
                        x = 1.13,
                        y = 1.05,
                        text = 'Magnitude',
                        xref = 'mag',
                        yref = 'mag',
                        showarrow = FALSE
                        ))
longlatdepth
```

Let's do some more 3D exploring. How do the errors look? And let's color them based on magnitude to see if the errors are correlated with the recorder magnitude


```{r}

longlatdepth <- plot_ly(df, x = ~horizontalError, y = ~depthError, z = ~magError,
               marker = list(color = ~mag, colorscale = c('#FFE1A1', '#683531'), showscale = TRUE))
longlatdepth <- fig %>% add_markers()
longlatdepth <- fig %>% layout(scene = list(xaxis = list(title = 'Horizontal Error'),
                                   yaxis = list(title = 'Depth Error'),
                                   zaxis = list(title = 'Magnitude Error')),
                      annotations = list(
                        x = 1.13,
                        y = 1.05,
                        text = 'Magnitude',
                        xref = 'mag',
                        yref = 'mag',
                        showarrow = FALSE
                        ))
longlatdepth
```

Ok, now let's take a look at some animations. (be patient when loading them)
Not extremely useful, but super cool to look at!

```{r}
ggplot(data = df) +
  geom_point(mapping = aes(x = longitude, y = latitude, color = mag)) +
  labs(title = 'Year: {frame_time}', x = 'Longitude', y = 'Latitude') +
  transition_time(yearInt) +
  ease_aes('linear')
```

```{r}
ggplot(data = df) +
  geom_bar(mapping = aes(x = mag), fill = 'steelblue') +
  labs(title = 'Year: {frame_time}', x = 'Mag') +
  transition_time(yearInt) +
  ease_aes('linear')
```


....and we tried this library as well, but, for some reason, it doesn't work. Too bad.

```{r}

gg_point = ggplot(data = df) +
    geom_point_interactive(aes(x = longitude, y = latitude, color = year)) + 
  theme_minimal()

```


```{r}
get_ts_matrix <- function(ts_magnitude) {
  time <- 0
  mag <- 0
  for (index in 1:nrow(ts_magnitude)) {
    time[index] <- ts_magnitude[index,1]
    mag[index] <- ts_magnitude[index,2]
  }
  mag_matrix <- matrix(ts_magnitude$df.mag, 
                       nrow =length(ts_magnitude$df.timeYear),
                       dimnames = list(ts_magnitude$df.timeYear, "data"))
  return(mag_matrix)
}
```










```{r}
### CHECK FOR STATIONARITY AND MAKE 
### DATA STATIONARY

### Magnitude time series

ts_magnitude <- data.frame(df$timeYear, df$mag)

ts_magnitude_plot <- ggplot(ts_magnitude, aes(x=df.timeYear, y=df.mag)) +
  ggtitle("Magnitude time series") +
  geom_line() + 
  xlab("")

ts_magnitude_plot

mag_matrix <- get_ts_matrix(ts_magnitude)
acf2(ts(mag_matrix))

### First diff time series
ts_magnitude$df.mag<- c(0, diff(df$mag))

ts_diff_magnitude_plot <- ggplot(ts_magnitude, aes(x=df.timeYear, y=df.mag)) +
  ggtitle("First diff - Magnitude time series") +
  geom_line() + 
  xlab("")

ts_diff_magnitude_plot

diff_mag_matrix <- get_ts_matrix(ts_magnitude)
acf2(ts(diff_mag_matrix))

### Second diff time series

### ts_magnitude$df.mag<- c(0, diff(ts_magnitude$df.mag))

ts_second_diff_magnitude_plot <- ggplot(ts_magnitude, aes(x=df.timeYear, y=df.mag)) +
  ggtitle("Second diff - Magnitude time series") +
  geom_line() + 
  xlab("")

ts_second_diff_magnitude_plot

second_diff_mag_matrix <- get_ts_matrix(ts_magnitude)
acf2(ts(second_diff_mag_matrix))

### ACF indicates q
### PACF indicates p
### arima(p,diff,q)

data <- ts(diff_mag_matrix)

ma1<-arima(data, order=c(0,0,1))

ar1<-arima(data ,order=c(1,0,0))
ar2<-arima(data ,order=c(2,0,0))
ar3<-arima(data ,order=c(3,0,0))
ar4<-arima(data ,order=c(4,0,0))
```

```{r}
my_model <- arima(data, order=c(1,2,2))
stdres <- my_model$residuals/sqrt(my_model$sigma2)
plot(my_model$residuals, main = "Chosen model - ARIMA(1, 3, 2) - Residuals")
qqnorm(stdres, main = "Chosen model - ARIMA(1, 3, 2) - Normal Q-Q Plot of Std Residuals")

shapiro.test(residuals(my_model))
```

```{r}
my_forecast = predict(arima(ts(mag_matrix), order=c(1,2,2)), n.ahead=30)
# ts.plot(ts(mag_matrix), 
#        my_forecast$pred,  
#        col=1:2,
#        lwd=1)

my_forecast
ggplot(ts_magnitude, aes(x=df.timeYear, y=df.mag)) +
  ggtitle("Magnitude time series") +
  geom_line() + 
  xlab("")
```
