---
title: "Info Visualisation Notebook"
output: info_vis_notebook
---

Set working directory

```{r}
# Ana's WD
# setwd("C:/Users/Ana-Maria's Laptop/Documents/GitHub/infoVisualization")

# Andrei's WD
# setwd("C:/Users/andre/Documents/infoVisualization/")
```

Imports - libraries and read dataset

```{r}
#install.packages("png")
#install.packages("gifsky")
#install.packages("gganimate")
#install.packages("gapminder")
#install.packages("ggmap")
#install.packages("esquisse")
#install.packages("ggiraph")
#install.packages("gdtools")
#install.packages("leaflet") # interactive maps



library(ggplot2)
library(gganimate)
library(gapminder)
#library(tidyverse)
library(ggmap)
library(ggiraph)

library(esquisse)
library(leaflet)


# esquisser(df)
```

```{r}
earthquakes_csv = "eartquakes_Romania.csv"
df <- read.csv(file = earthquakes_csv)
```

```{r}
# convert time column to R timestap
df$time <- gsub('T', ' ', df$time)
df$time <- gsub('Z', '', df$time)
df$time <- as.POSIXct(df$time, format="%Y-%m-%d %H:%M:%OS", tz=Sys.timezone())

#split timestamp into date and time
df$timeYear <- as.Date(df$time)
df$timeDay <- format(df$time, "%H:%M:%OS")

df$year <- format(df$time, "%Y")
df$month <- format(df$time, "%m")

df$yearInt <- as.integer(df$year)
```

```{r}
# see data types
str(df)
```

```{r}
# Exploring values to better understand the dataset
# Starting with mag (magnitude)
mag_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = mag), fill = "steelblue", bins=40) +
  #geom_freqpoly(mapping = aes(x = mag), bins=40) +
  theme_minimal()

mag_hist + ggtitle("Number of recorded earthquakes by magnitude")

# let's see how many are per year
year_bar <- ggplot(df) + 
  geom_bar(mapping = aes(y = year), fill = "steelblue") +
  theme_minimal()

year_bar + ggtitle("Number of earthquakes per year (1990 - 2017)")

# and per month
month_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = month), fill = "steelblue") +
  theme_minimal()

month_bar + ggtitle("Number of earthquakes by month (1990 - 2017)")

# earthquakes count by depth

depth_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = depth), fill = "steelblue") +
  theme_minimal()

depth_bar + ggtitle("Number of earthquakes by depth")

# magtype count

magtype_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = magType), fill = "steelblue") +
  theme_minimal()

magtype_bar + ggtitle("Magnitude type (?)")


# nst count

nst_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = nst), fill = "steelblue", bins=50) +
  theme_minimal()

nst_bar + ggtitle("Nst type (?)")



# gap
gap_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = gap), fill = "steelblue") +
  theme_minimal()

gap_bar + ggtitle("Gap Value (?)")



# dmin
dmin_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = dmin), fill = "steelblue") +
  theme_minimal()

dmin_bar + ggtitle("Dmin Value (?)")

# rms
rms_bar <- ggplot(df) + 
  geom_histogram(mapping = aes(x = rms), fill = "steelblue") +
  theme_minimal()

rms_bar + ggtitle("Rms Value (?)")

# net
net_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = net), fill = "steelblue") +
  theme_minimal()

net_bar + ggtitle("Net Value (?)")


# type
type_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = type), fill = "steelblue") +
  theme_minimal()

type_bar + ggtitle("Type Value (?)")


# status
status_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = status), fill = "steelblue") +
  theme_minimal()

status_bar + ggtitle("Status Value (?)")


# LocationSource
locationSource_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = locationSource), fill = "steelblue") +
  theme_minimal()

locationSource_bar + ggtitle("Location Source Value (?)")


# MagSource
magSourceSource_bar <- ggplot(df) + 
  geom_bar(mapping = aes(x = magSource), fill = "steelblue") +
  theme_minimal()

magSourceSource_bar + ggtitle("Mag Source Value (?)")



# horizontalError
horizontalError_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = horizontalError), fill = "steelblue") +
  theme_minimal()

horizontalError_hist + ggtitle("horizontalError Value (?)")


depthError_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = depthError), fill = "steelblue") +
  theme_minimal()

depthError_hist + ggtitle("depthError Value (?)")

magError_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = magError), fill = "steelblue") +
  theme_minimal()

magError_hist + ggtitle("magError Value (?)")



magNst_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = magNst), fill = "steelblue") +
  theme_minimal()

magNst_hist + ggtitle("magNst Value (?)")



mag_hist <- ggplot(df) + 
  geom_histogram(mapping = aes(x = mag), fill = "steelblue", bins = 40) +
  #geom_freqpoly(mapping = aes(x = mag), bins=40) +
  theme_minimal()


```

```{r}
# We start by exploring the most obvious questions: where do most earthquakes happen, including their magnitude as well

ggplot(data=df) +
  geom_point(mapping = aes(x = longitude, y = latitude, color=mag))

# as we can see, there is a big cluster where most earthquakes happen, with a smaller cluster that is noticeable, followed by sparsely distributed cases


ggplot(data=df) +
  geom_bar(mapping = aes(x = mag, fill = year))


ggplot(data=df) +
  geom_freqpoly(mapping = aes(x = mag, color = year))

ggplot(data=df) +
  geom_density(mapping = aes(x = mag, color = year))


#ggplot(data=df) +
#  geom_point(mapping = aes(x = mag, y = year))
```

```{r}
# Visualizing the earthquakes and their magnitudes each year 
ggplot(df) +
  geom_bar(mapping = aes(x = mag, fill = year)) +
  facet_wrap(~ year)

```

`heme_minimal()`

```{r}
get_ts_matrix <- function(ts_magnitude) {
  time <- 0
  mag <- 0
  for (index in 1:nrow(ts_magnitude)) {
    time[index] <- ts_magnitude[index,1]
    mag[index] <- ts_magnitude[index,2]
  }
  mag_matrix <- matrix(ts_magnitude$df.mag, 
                       nrow =length(ts_magnitude$df.timeYear),
                       dimnames = list(ts_magnitude$df.timeYear, "data"))
  return(mag_matrix)
}
```

```{r}
ggplot(data=df) +
  geom_point(mapping = aes(x = longitude, y = latitude, color = mag)) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'Longitude', y = 'Latitude') +
  transition_time(yearInt) +
  ease_aes('linear')
```

```{r}
ggplot(data=df) +
  geom_bar(mapping = aes(x = mag)) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'Mag') +
  transition_time(yearInt) +
  ease_aes('linear')
```

```{r}
romania_map <- get_map("Brasov", maptype = "roadmap", zoom = 6)

romania_box <- c(20, 43, 30, 48.5)
romania_map2 <- get_map(location = romania_box, source = "stamen", maptype = "toner", zoom = 7)


#ggmap(romania_map) + 
#  geom_point(data = df, aes(x = longitude, y = latitude), color = "steelblue", size = 1) 

ggmap(romania_map2) + 
  geom_point(data = df, aes(x = longitude, y = latitude, color = mag), size = 1) 

```

```{r}
gg_point = ggplot(data = mtcars) +
    geom_point_interactive(aes(x = wt, y = qsec, color = mag, tooltip = carname, data_id = carname)) + 
  theme_minimal()

gg_point = ggplot(data = df) +
    geom_point_interactive(aes(x = longitude, y = latitude, color = year)) + 
  theme_minimal()

```

getColor <- function(quakes) {
  sapply(quakes$mag, function(mag) {
  if(mag <= 4) {
    "green"
  } else if(mag <= 5) {
    "orange"
  } else {
    "red"
  } })
}

icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'black',
  library = 'ion',
  markerColor = getColor(df)
)
```

```{r}
# Show first 20 rows from the `quakes` dataset
leaflet(data = df) %>% 
  addTiles() %>%
  addAwesomeMarkers(~longitude, ~latitude, popup = ~as.character(year), label = ~as.character(mag), icon=icons, clusterOptions = markerClusterOptions())
```

```{r}
library(leaflet)

m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=174.768, lat=-36.852, popup="The birthplace of R")
m  # Print the map
```

```{r}
### CHECK FOR STATIONARITY AND MAKE 
### DATA STATIONARY

### Magnitude time series

ts_magnitude <- data.frame(df$timeYear, df$mag)

ts_magnitude_plot <- ggplot(ts_magnitude, aes(x=df.timeYear, y=df.mag)) +
  ggtitle("Magnitude time series") +
  geom_line() + 
  xlab("")

ts_magnitude_plot

mag_matrix <- get_ts_matrix(ts_magnitude)
acf2(ts(mag_matrix))

### First diff time series
ts_magnitude$df.mag<- c(0, diff(df$mag))

ts_diff_magnitude_plot <- ggplot(ts_magnitude, aes(x=df.timeYear, y=df.mag)) +
  ggtitle("First diff - Magnitude time series") +
  geom_line() + 
  xlab("")

ts_diff_magnitude_plot

diff_mag_matrix <- get_ts_matrix(ts_magnitude)
acf2(ts(diff_mag_matrix))

### Second diff time series

### ts_magnitude$df.mag<- c(0, diff(ts_magnitude$df.mag))

ts_second_diff_magnitude_plot <- ggplot(ts_magnitude, aes(x=df.timeYear, y=df.mag)) +
  ggtitle("Second diff - Magnitude time series") +
  geom_line() + 
  xlab("")

ts_second_diff_magnitude_plot

second_diff_mag_matrix <- get_ts_matrix(ts_magnitude)
acf2(ts(second_diff_mag_matrix))

### ACF indicates q
### PACF indicates p
### arima(p,diff,q)

data <- ts(diff_mag_matrix)

ma1<-arima(data, order=c(0,0,1))

ar1<-arima(data ,order=c(1,0,0))
ar2<-arima(data ,order=c(2,0,0))
ar3<-arima(data ,order=c(3,0,0))
ar4<-arima(data ,order=c(4,0,0))
```

```{r}
my_model <- arima(data, order=c(1,2,2))
stdres <- my_model$residuals/sqrt(my_model$sigma2)
plot(my_model$residuals, main = "Chosen model - ARIMA(1, 3, 2) - Residuals")
qqnorm(stdres, main = "Chosen model - ARIMA(1, 3, 2) - Normal Q-Q Plot of Std Residuals")

shapiro.test(residuals(my_model))
```

```{r}
my_forecast = predict(arima(ts(mag_matrix), order=c(1,2,2)), n.ahead=30)
# ts.plot(ts(mag_matrix), 
#        my_forecast$pred,  
#        col=1:2,
#        lwd=1)

my_forecast
ggplot(ts_magnitude, aes(x=df.timeYear, y=df.mag)) +
  ggtitle("Magnitude time series") +
  geom_line() + 
  xlab("")
```
